
\begin{abstract}
    Lens assembly design is a complex and time-intensive process that requires navigating a high-dimensional design space while satisfying strict performance and feasibility constraints. Modern commercial optical design software alleviates part of this burden by enabling local optimization of well-initialized designs. However, identifying high-quality starting points remains a non-trivial and often manual task, and strongly influences the final design outcome. Deep generative models have shown promise for design space exploration in several engineering domains, but their adoption in optical system design has been limited. And, in practice, candidate designs are still largely generated through expert-driven, time-consuming workflows. To address these limitations, we present: 
    (1) A large scale dataset of patented and synthetically expanded lens assembly designs and their performance metrics; (2) A parametric representation for optical systems that significantly improves generative model performance; (3) A ray-tracing-based post-processing pipeline that significantly improves feasible sample fraction and performance metrics of generative model samples (4) A systematic benchmarking study using off-the-shelf unconditional generative models; and (5) a conditional latent diffusion model architecture that can produce physically meaningful designs conditioned on design requirements. The proposed conditional generative model architecture consists of a transformer-based autoencoder that embeds variable-length assemblies into a fixed-length latent space and a conditional diffusion model trained to generate latent representations for specified design conditions. Together with the novel representation and post-processing pipeline, these components enable condition-guided exploration of lens assembly design spaces and provide initialization points for downstream optimization tasks.
\end{abstract}



